---
title: "gofl Design"
author: "Bradley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gofl design}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# The Language

`gofl` is built around 3 algebraic operators:

* `+`: the sum operator
* `:`: the product operator
* `*`: the sum of the sum and product

By design, these operators correspond to the same operators in `R` model formulas. Under the hood, these operators are replaced in the AST by internal generic operators `%s%`, `%p%`, and `%ssp%` 

Additionally, `gofl` has two functions, `tag` and `.zoom`, which have specific use cases.

# The Algebra

The work of `gofl` is done by defining how sum (`%s%`) and product (`%p%`) work on different types.

Permuation matrices are used to define variable levels:

```{r}
X <- gofl:::as_tmatrix("var1", c("level1", "level2", "level3"))
Y <- gofl:::as_tmatrix("var2", LETTERS[1:5])
```

Each row corresponds to a group:

```{r}
X@mat
Y@mat
```

The sum of these two matrices is the direct sum, or block diagonal matrix:

```{r}
gofl:::`%s%`(X@mat, Y@mat)
```

The product of these two matrices is the cartesian product.

```{r}
gofl:::`%p%`(X@mat, Y@mat)
```

The sum of the sum and product does both the above operations but columns with the same name are stacked appropriately: 

```{r}
gofl:::`%ssp%`(X@mat, Y@mat)
```

These operators are defined analoguouly are defined for lists. The sum of two lists is concatenation:

```{r}
gofl:::`%s%`(list(1, 2), list(2, 3, 5))
```

The product of two lists is the cartesian product:

```{r}
gofl:::`%p%`(list(1, 2), list(2, 3, 5))
```

The same operators also work for integers:

```{r}
gofl:::`%s%`(2L, 3L)
gofl:::`%p%`(2L, 3L)
gofl:::`%ssp%`(2L, 3L)
```

To get `gofl` to work a data structure called `tagged` holds the matrix and the tags:

```{r}
str(X)
```

Then the operators are applied simultaneously to the matrices and the tags.
```{r}
gofl:::`%s%`(X, Y)
```

# Hijacking the AST

The `traverse_expr` function replaces `+` with `%s%` and so on.

```{r}
ff <- ~ a*b + c:d + e
nf <- gofl:::traverse_expr(ff, f = identity)
# It's not pretty to look at
nf
```

Now the new expression can be evaluated with data. Usually, our `data` here would be the `tagged` objects, but just for illustration:

```{r}
rlang::eval_tidy(nf[[2]], data = list(a = 1L, b = 3L, c = 4L, d = 2L, e =  7L))
```

The `traverse_expr` function can also take a function as an argument that is applied to each leaf of the AST. In this example `gofl:::replace_by_size` finds the `nrow` of a matrix.

```{r}
rlang::eval_tidy(
  expr = gofl:::traverse_expr(ff, f = gofl:::replace_by_size)[[2]], 
  data = list(
  a = matrix(nrow = 1), 
  b = matrix(nrow = 3), 
  c = matrix(nrow = 4), 
  d = matrix(nrow = 2),
  e = matrix(nrow = 7)))
```

To continue the example from above and demonstrate how it works on the `tagged` type:

```{r}
ff <- ~ var1 + var2
rlang::eval_tidy(
  expr = gofl:::traverse_expr(ff, identity)[[2]], 
  data = list(var1 = X, var2 = Y))
```
