---
title: "Using gofl"
author: "Bradley Saul"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using gofl}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = TRUE, message=FALSE, warning=FALSE}
library(gofl)
library(dplyr)
```

# Motivation

A common task in data science pipelines is to summarize a dataset by subgroups. For example, consider everyone's (least) favorite dataset `mtcars`:

```{r}
str(mtcars)
```

This dataset contains two binary variables by which one may be interested in grouping the data: `vs` and `am`. Two binary variables present 3 grouping (for $2^3 = 8$ possible groups total) for summarizing the data: marginally by the two groups within each variable, or jointly by the 4 groups defined by the cross of the two variables.

One way to obtain these groups is to use `dplyr::group_by`. By way of example, say we are interested in the mean miles per gallon within each of these groups:

```{r}
summarize_mpg <- function(dt) summarize(dt, mean(mpg))

mtcars %>%
  group_by(vs) %>%
  summarize_mpg

mtcars %>%
  group_by(am) %>%
  summarize_mpg

mtcars %>%
  group_by(vs, am) %>%
  summarize_mpg
```
   
That works, but as the number of grouping variables increases, this approach becomes untenable. `gofl` takes a different approach. The following `gofl` formula specifies groupings for each of the margins and the joint:

```{r}
mt_groups <- ~ vs + am + vs:am
# or equivalently
mt_groups <- ~ vs*am
```

A `gofl` specification is two parts: a `formula` defining the groupings and a `list` defining levels of the variables in the formula. At this time, the `list` of levels needs to be created by hand. Future versions of `gofl` may provide a means to automate this step. 

```{r}
dat <- list(
 vs = sort(unique(mtcars$vs)), # The sorting isn't strictly necessary.
 am = sort(unique(mtcars$am))
)
```

With these two pieces of data the grouping specification can be created:
```{r}
grps <- create_groupings(formula = mt_groups, data = dat)
```

The result of `create_groupings` is a list with three elements: `r paste(names(grps), collapse = ", ")`. The `data` element is a just copy of the `data` argument. The `groupings` element is a `list` where each element is a specification for each group defined by the formula. In our example, the first group defines the subset containing `vs == 0`:

```{r}
str(grps$groupings[[1]])
```

A single group's specification has 4 parts:

* `i`: the index of the group (described below)
* `q`: a `quosure` that can be used to carry out the subgrouping in (e.g.) `dplyr::filter`
* `g`: a `list` that specifies which variables and levels define the group
* `tags`: a `character` vector of tags (See tags section)

Here's another example of a grouping; in this case it is the subset where `vs == 1` and `am == 0`:

```{r}
str(grps$groupings[[7]])
```

## Using grouping soecification to obtain summaries

With the groupings now defined we can now create a pipeline for summarizing within each group.

```{r}
purrr::map_dfr(
  .x = grps$groupings,
  .f = ~ {
    mtcars %>% 
    filter(!!! .x$q) %>% 
    summarize_mpg() %>%
    bind_cols(tibble(!!! .x$g))
  }
)
```

That's the basic idea of `gofl`. From the patterns described above you can specify much more complicated groupings.

# Indices

Each element of a grouping is even a unique index as follows:

* each variable is given a positional index in the pattern of `v1-v2-v3-...-vn`, where the order is the order of the variables given in the `create_groupings` `data` argument. In the example, above `vs` corresponds to the first position (`v1`) and `am` to the second (`v2`).
* each level of a variable is given a positive integer value. In the example, `vs = 0` corresponds to `1`, `vs = 1` corresponds to `2`. A group not involving `vs` corresponds to `0`.

The index `1-0` in our example is the group defined by `vs == 0`; `0-2` is the group defined by `am == 1`; `1-2` is defined by `vs == 0 & am == 1`. 

The `index_fcn` part of a `create_groupings` object allows the user to look up indices.

```{r}
grps$index_fcn(am = 0)
grps$index_fcn(am = 0, vs = 1)
grps$index_fcn(vs = 1)
```

Indices are useful when the number of groups is large and you need a way to quickly find a particular group.

# Tagging

In real applications, you may want downstream processes to apply different functions to different groups. The `tag` function allows you to tag particular groups by arbitrary `character` vectors. Let's make our example above even more unrealistic but nonetheless illustrative and take the mean in the marginal groups and the median in joint group.  

First, we tag the grouping as appropriate:
```{r}
mt_groups <- ~ tag(vs + am, "marginal") + tag(vs:am, "joint")
grps <- create_groupings(formula = mt_groups, data = dat)
```

Note that the `tag` element is now populated:

```{r}
grps$groupings[c(1,5)]
```

An example pipeline taking advantage of the `tag` may look like:

```{r}
summarize_mpg2 <- function(dt, tag){
  avg <- switch(tag, "marginal" = mean, "joint" = median)
  summarize(dt, avg(mpg))
}

purrr::map_dfr(
  .x = grps$groupings,
  .f = ~ {
    mtcars %>% 
    filter(!!! .x$q) %>% 
    summarize_mpg2(.x$tag) %>%
    bind_cols(tibble(!!! .x$g))
  }
)
```
